---
title: "Project_2_Section_3"
author: "Maryam"
date: "2025-04-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Section 3: Data Cleaning

This section includes steps to identify and fix data entry errors and transform problematic variables in the King County housing data-set.

```{r}
library(tidyverse)
library(scales)
library(gridExtra)
```

```{r}
#load and inspect data
Data <- read.csv("kc_house_data.csv", header = TRUE)

```

### 3.1 Transform date Column

The date column is a character string and not useful for modeling. We'll extract: year and month sold.

```{r}

Data$date <- as.character(Data$date)

Data$year_sold <- as.numeric(substr(Data$date, 1, 4))
Data$month_sold <- as.numeric(substr(Data$date, 5, 6))

table(Data$year_sold)
table(Data$month_sold)
```
### 3.2 Transform zipcode to a Factor

Zipcodes are stored as a numeric variables, but they aren’t meant to be used like regular numbers. They're more like location labels, not values. It needs to be converted to a factor so R treats it as category.


```{r}

Data$zipcode <- as.factor(Data$zipcode)

str(Data$zipcode)
length(unique(Data$zipcode))
```

There were 70 unique zipcodes in the dataset.


### 3.3 Converting Renovation Year to a Binary Renovation (Yes/No)

```{r}

Data$renovated <- ifelse(Data$yr_renovated > 0, 1, 0)
table(Data$renovated)
```

# SECTION 4 - VISUALIZATIONS

```{r}
beds <- ggplot(Data, aes(x=bedrooms, y=price))+
  scale_y_continuous(breaks = breaks_extended(6),labels = label_dollar())+
  stat_summary(geom="bar")

baths <- ggplot(Data, aes(x=bathrooms, y=price))+
  scale_y_continuous(breaks = breaks_extended(6),labels = label_dollar())+
  stat_summary(geom="bar")

grid.arrange(beds, baths, ncol=2,nrow=1)
```

```{r}
ggplot(Data, aes(x=sqft_living, y=price))+
  scale_y_continuous(breaks = breaks_extended(6),labels = label_dollar())+
  geom_point()
```

```{r}
ggplot(Data, aes(x=sqft_lot, y=price))+
  scale_y_continuous(breaks = breaks_extended(6),labels = label_dollar())+
  geom_point()
```


```{r}
ggplot(Data, aes(x=grade, y=price))+
  scale_y_continuous(breaks = breaks_extended(6),labels = label_dollar())+
  geom_point()
```
```{r}
grade <- ggplot(Data, aes(x=grade, y=price))+
  scale_y_continuous(breaks = breaks_extended(6),labels = label_dollar())+
  stat_summary(geom="bar")

condition <- ggplot(Data, aes(x=condition, y=price))+
  scale_y_continuous(breaks = breaks_extended(6),labels = label_dollar())+
  stat_summary(geom="bar")

grid.arrange(grade,condition,ncol=2,nrow=2)
```

```{r}
ggplot(Data, aes(x=month_sold))+
  geom_density()
```



```{r}
write.csv(Data, "kc_house_data_cleaned.csv", row.names = FALSE)

```

